---
title: "Deliverable 3 Data Analyses"
output: html_document
date: "2024-03-14"
authors: Anderson,Cabe,Gabby
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

## ANOVA and graphs comparing season and mosquito pathogen positivity across sites

```{r, message = FALSE}
# Loading appropriate packages

library(tidyverse)
library(here)
library(cowplot)
library(viridis)
```

```{r}
# Read in data
temp_day <- read_csv(here("data", "processed", "temp_day.csv"))
mosquito_clean <- read_csv(here("data", "processed", "mosquito_clean.csv"))
```

```{r}
#Convert mosquito test data into binary values
mosquito_binary <- mosquito_clean %>% 
  mutate(testResult_mos= ifelse(testResult=="Positive",1,0),
         collectDate = floor_date(endCollectDate, "day")) %>% 
  select(-startCollectDate, -endCollectDate)
```

```{r}
#Saving positive/negative binary data to processed data folder
mosquito_binary %>% 
  write_csv(here("data", "processed", "mosquito_binary.csv"))
```

```{r}
#Joining data sheets, making a new season column and dropping unneeded variables
mosquito_day <- mosquito_clean %>% 
  select(-startCollectDate) %>% 
  mutate(collectDate = as_date(endCollectDate)) %>% 
  select(-endCollectDate) 

temp_day_join <-temp_day %>% 
  rename("collectDate"= `floor_date(startDateTime, "day")`)

temp_mos_raw <-
  full_join(mosquito_binary, temp_day_join) %>% 
  select(-testResult)

temp_mos_raw #%>% 
  #write_csv(here("data", "processed", "temp_mos_raw.csv"))

temp_mos_season <- temp_mos_raw %>% 
  drop_na() %>% 
  subset %>% 
  mutate(month = month(collectDate),
         season = ifelse(month %in% c(06,07,08), "summer",
                         ifelse(month %in% c(09,10,11), "fall",
                                ifelse(month %in% c(12,01,02), "winter",
                                        ifelse(month %in% c(03,04,05), "spring",
                                                "NA"))))) %>% 
  group_by(siteID, season, testResult_mos) %>% 
  summarize(total = n())

temp_mos_season

```

```{r}
#Performing an ANOVA on test results with season and site id
model1 <- aov(testResult_mos ~ siteID * season, data = temp_mos_season)

summary(model1)
TukeyHSD(model1)

par(mfrow = c(1, 2))
plot(model1, which = c(2, 3))

#Additive model
additive_model <- aov(testResult_mos ~ siteID + season, data = temp_mos_season)

summary(additive_model)
TukeyHSD(additive_model)

par(mfrow = c(1, 2))
plot(additive_model, which = c(2, 3))

#We performed a two-way ANOVA to test whether the proportion of positive 
#mosquito test results were not associated with the site ID, the season of the year, 
#or the interaction of the two. None of the variables listed above had a 
#significant relationship (all p values > 0.85).
```

```{r}
#Graph of proportion of positive results over seasons across sites
temp_mos_season_prop <- temp_mos_season %>% 
  pivot_wider(names_from = testResult_mos,
              values_from = total,
              values_fill = 0) %>% 
  mutate(prop_pos = `1` / sum(`0` + `1`)) %>% 
  mutate(season = factor(season,levels = c("fall", "winter", "spring", "summer")))

ggplot(data = temp_mos_season_prop, 
         aes(x = season, 
             y = prop_pos,
             col = siteID)) +
    geom_jitter(cex = 3,
                height = 0,
                width = 0.1) +
  labs( x = "Season", y = "Proportion of positive test results") +
  theme_bw()
  
```

# Proportion of viral disease presence in mosquitoes separated by season of the
# year and NEON site from which data was gathered (DSNY = Disney World, Orlando, Florida; HARV = Cambridge, MA; MOAB = Moab, Utah; WOOD = Woods Hole, 
# Massachussetts. The only positive test results occurred in summer, with Woods
# Hole, MA, having the highest proportion of positives.

```{r}
#Graph of proportion of positive results of virus types across sites 
temp_mos_pathogen <- temp_mos_raw %>% 
  drop_na() %>% 
  group_by(siteID, testResult_mos, testPathogenName) %>% 
  summarize(total = n()) %>% 
  pivot_wider(names_from = testResult_mos,
              values_from = total,
              values_fill = 0) %>% 
  mutate(prop_pos = `1` / sum(`0` + `1`))

ggplot(data = temp_mos_pathogen, 
         aes(x = siteID, 
             y = prop_pos,
             col = testPathogenName)) +
    geom_jitter(cex = 3,
                height = 0,
                width = 0.1) +
  labs( x = "Site", y = "Proportion of positive test results", color = "Virus") +
  theme_bw()

```
# Proportion of positive test results separated by site ID and pathogen. 
# The only pathogens detected were Flavivirus and West Nile, and the prevalence
# of both viruses was higher in Woods Hole than Disney.

