---
title: "Accessing NEON raw data"
author: "Your Name"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
date: "YYYY-MM-DD"
---

Using R to pull NEON data directly from the [Data Portal website](https://data.neonscience.org/data-products/explore) allows us to avoid the frustration of unzipping nested folders and having to manually combine multiple months and/or sites of data into a single file to analyze. 

Pulling our data via scripts also makes our research process more transparent and reproducible!

The first step to get ready to "grab" NEON data is to install the `neonUtilities` package, using the code chunk below. We will also install the `here` package, which we will use later on when saving our raw data:

```{r load libraries, message = FALSE}

### Un-hash the following two lines to run the "install" commands the first time you run this chunk, then re-hash both lines to avoid problems knitting your Markdown document

#install.packages("neonUtilities")
#install.packages("here")

library(tidyverse)
library(neonUtilities)
library(here)

```

## Specify NEON data products & sites

For each type of [NEON data](https://data.neonscience.org/) you want to download from the Portal, you will need to specify:

-   The specific data product ID (e.g., "DP1.10086.001")

-   The 4-letter code for your target NEON sites (e.g., "HARV"). Omit to download all available dates (resulting file may be *large*)

-   The YYYY-MM start and end dates (e.g., "2019-12"). Omit to download all available dates (resulting file may be *large*).

It will likely be helpful to visit the [Data Portal](https://data.neonscience.org/data-products/explore) through a web browser to find your product ID and target site codes.

Update the code chunk below to specify your target data product(s) and site(s). If you want to pull all sites and available dates, you can leave those lines blank.
```{r}

# Hantavirus rodent pathogen status
Rodent_pathogen_status_hantavirus <- "DP1.10064.001"

# Two selected focal sites
my_sites <- c("DSNY", "HARV", "WOOD", "MOAB")

# Start date (as YYYY-MM)
start_date <- "2019-01"

# End date (as YYYY-MM)
end_date <- "2022-10"

```

## Set up data-grab commands: 
When using the `loadByProduct()` command within neonUtilities, you are only able to download one data package at a time.  

NEON data are saved to R as a *list* of data files by default. 
We use the `list2env()` command as a wrapper so that data files are each saved as separate objects in our RStudio environment.

After you initiate the following command, you will see the following in the Console: 

> Finding available files ================================================| 100%  
> Rate limit reached. Pausing for 99 seconds to reset. **(or not, if your file is small)**  
> Continuing will download files totaling approximately 5.578909 MB.  
> Do you want to proceed y/n: 
  
**You will need to type y in the Console to proceed with the download**

```{r}

# We'll pull the rodent data first
list2env((loadByProduct(dpID = Rodent_pathogen_status_hantavirus, 
                        site = my_sites, 
                        include.provisional = TRUE,
                        startdate = start_date,
                        enddate = end_date)),
         .GlobalEnv)

```
## Save raw data files:
First, we'll save a "full" raw version of the relevant data files that we can use as a starting point for future analyses. This is important so we do not need to go through the download steps every time we are working on our project!

To ensure your work is reproducible, remember to *always* save a copy of the metadata file ("variables" file within NEON downloads) within your `raw` data folder.  

You should **also** access the data product on the [NEON Data Portal](https://data.neonscience.org/data-products/explore) and save a copy of the "Quick Start Guide" to your `raw` data folder as a .PDF or .md file.

```{r save metadata csv}

# Variable names and units: rodents
variables_10064 %>% 
  write_csv(here("data", "raw", "variables_10064.csv"))

```
Since our beetle count data would be in the "bet_sorting" table, we want to save that file to our `raw` data subfolder. It is likely also prudent to go ahead and save the bet_fielddata, since that provides information about the sampling sites. 

Remember, if you do *not* initially save a raw file but realize you need it later, you can always re-run the NEON data download commands for the same site(s) and date range to save additional files. 

```{r save raw data from relevant data tables}

# bet_sorting
rpt_bloodtesting %>% 
  write_csv(here("data", "raw", "rpt_bloodtesting.csv"))


```

## Download subsequent data product(s)
We'll repeat the commands above to download and save the raw data for our second data product of interest (precipitation).

Save individual raw data files of interest (here, throughfull precipitation every 30 minutes):


Because the downloads can occupy a lot of space on our computer (& in our global environment), we'll often do a "clear" command once we've saved our raw data. 

```{r}

# Clear global environment
rm(list = ls())

```


We would then start our data exploration/cleaning in a separate .Rmd file that begins with:
* Loading the `tidyverse` and `here` libraries again  
* Using a `read_csv()` command to import our raw data files from the `raw` subfolder  